ASAH1 rGBM Manuscript ELISA Figure

### 2023-02-01 results/elisa_plasma_proteomics/manuscript_elisa_figure_analysis.Rmd
[RENAME] notebook to analysis, since more exploratory analyses were included

<!-- # Helper Functions -->
```{r message=FALSE, warning=FALSE, include=FALSE}
#' @title Master Trainer and Prediction
#' @description Top Function to perform predictions for testing logistic regression, logistic regression + elastic regularization, random forest and k-Nearest Neightbours
#' 
#' @param train_df (data.frame) A dataframe that contains labels and features to run through prediction models. Has to have the following structure:
#' \item{class}{(character) sample/observation label. i.e. ctrl, iGBM, or rGBM sample.}
#' \item{feature_1}{(numeric} a specific feature measurement such as protein ASAH1)}
#' \item{feature_n}{(numeric) another specific feature.}
#' @param pred_col_prob (character) A character corresponding to the class of probabilities you want to extract for. i.e. "GBM"
#' @param multiclass (logical) Run predictions using multiclass modelling and prediction if more than 2 classes.
#' 
#' @return
#' 
#' returns a list object that contains trained model objects, lift data and roc data.
#' 
#' \item(model){(glmnet) Logistic Regression with Elastic Net Regularization Model }
#' \item(model_rf){(rf) Random Forest Model}
#' \item(model_logit){(glm) Logistic Regression Model}
#' \item(model_boostedlogit){(knn) k-Nearest Neighbours Models}
#' \item(master_lift){(data.table) lift data for all models}
#' \item(roc_data){(data.table) roc data for all models}
run_predictions <- function(train_df, pred_col_prob="GBM", multiclass=FALSE){

@@

#' Calculate the ratio of iGBM to rGBM and rGBM to iGBM for a given data frame
#'
#' @param data_long2 A data frame with columns "gene", "Number.x", "patient_state", "Age at surgery (years)", "Sex", "Tumor localization", "Type of resection", "MGMT promotor methylation", "p53 accumulation", "Ki67-Li", "EGFR vIII", "Treatment", "Time-to-reccurence (days)", "Latency (days)", and "conc".
#' @return A data frame with the calculated ratios.
get_ratio_data <- function(data_long2){
  ## Data wrangling ----
  ### iGBM over rGBM ratio
  data_long2 %>% 
    dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
    dplyr::group_by(gene, Number.x) %>%
    dplyr::group_map(function(.d, ...){
      .d <- as.data.table(.d)
      .d %>% 
        dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>% 
        dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                         agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
      # print(.d)
      if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
        .d_initial <- .d[manifestation_type=="Initial manifestation"]
        .d_second <- .d[manifestation_type=="Second manifestation"]
        tmp$ratio <- .d_initial$conc / .d_second$conc
      } else if (nrow(.d)==2) {
        .d_initial <- .d[collection_type=="E1"]
        .d_second <- .d[collection_type=="E2"]
        tmp$ratio <- .d_initial$conc / .d_second$conc
      } else {
        tmp$ratio <- NA #.d$conc
      }
      return(tmp)
    }, .keep = TRUE) %>%
    rbindlist() %>%
    dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
    dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
    dplyr::mutate(log2_conc = log2(ratio)) %>%
    # dplyr::group_by(gene, Number.x) %>%
    # dplyr::add_count() %>%
    # dplyr::filter(n==2) %>%
    dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata

  ### rGBM over iGBM
  data_long2 %>% 
    dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
    dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
    dplyr::group_by(gene, Number.x) %>%
    dplyr::group_map(function(.d, ...){
      .d <- as.data.table(.d)
      .d %>% 
        dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>% 
        dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                         agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
      # print(.d)
      if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
        .d_initial <- .d[manifestation_type=="Initial manifestation"]
        .d_second <- .d[manifestation_type=="Second manifestation"]
        tmp$ratio <- .d_second$conc / .d_initial$conc
      } else if (nrow(.d)==2) {
        .d_initial <- .d[collection_type=="E1"]
        .d_second <- .d[collection_type=="E2"]
        tmp$ratio <- .d_initial$conc / .d_second$conc
      } else {
        tmp$ratio <- NA #.d$conc
      }
      return(tmp)
    }, .keep = TRUE) %>%
    rbindlist() %>%
    dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "pre-surgery/post-surgery"))) %>%
    dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
    dplyr::filter(grepl(";", agg_lab_id)) %>%
    dplyr::mutate(log2_conc = log2(ratio)) %>%
    # dplyr::group_by(gene, Number.x) %>%
    # dplyr::add_count() %>%
    # dplyr::filter(n==2) %>%
    dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2

@@

@@

#' Calculate leave-one-patient-out cross-validation performance for splsda model
#'
#' This function takes in a dataset and a vector of patient IDs and returns a proportion
#' of correct predictions using a leave-one-patient-out cross-validation approach. A splsda
#' model is fit using the train data and used to make predictions on the test data. The
#' `unique_random_train_patient_per_class` parameter can be used to specify whether to 
#' include a unique patient for each "iGBM" and "rGBM" class in the train dataset. The
#' function returns a proportion of correct predictions.
#'
#' @param ratio_data A data frame containing the ratio data and class labels.
#' @param patient_ids A vector of patient IDs to use for the leave-one-patient-out
#' cross-validation.
#' @param unique_random_train_patient_per_class A logical value indicating whether to include
#' a unique patient for each "iGBM" and "rGBM" class in the train dataset.
#' @return A proportion of correct predictions.
#' @examples
#' splsda_loocv_perf(ratio_data, patient_ids)
splsda_loocv_perf <- function(ratio_data, patient_ids, unique_random_train_patient_per_class=TRUE){
  # Vector to store accuracy of LOOCV
  acc_vec <- numeric(length(patient_ids))
  i <- 1
  for (test_patient_id in patient_ids){
    # Split data into train and testing
    train_test <- get_train_test_for_loocv(ratio_data, test_patient_id=test_patient_id, unique_random_train_patient_per_class=unique_random_train_patient_per_class)
    train_df <- train_test$train_df
    test_df <- train_test$test_df
    # Perform splsda
    MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7, -8, -9, -10)], train_df$class, keepX = c(4,4)) 
    # Make prediction
    pred_test <- predict(MyResult.splsda, test_df[, c(-1, -2, -7, -8, -9, -10)], test_df$class)
    # evaluate the prediction accuracy for the first two components
    predict.comp2 <- pred_test$class$mahalanobis.dist[,2]
    # compute confusion matrix
    conf_mat <- table(factor(predict.comp2, levels = levels(train_df$class)), test_df$class)
    # print(conf_mat)
    # compute accuracy
    acc <- (conf_mat[1,1] + conf_mat[2,2])/(conf_mat[1,1] + conf_mat[2,2] + conf_mat[1,2] + conf_mat[2,1])
    # message(sprintf("INFO: Test Patient %s - predicted class = %s - model accuracy = %s", test_patient_id, predict.comp2, acc))
    # Store accuracy
    acc_vec[i] <- acc
    # update index
    i <- i + 1
  }
  # Compute proportion of correct predictions
  mean(acc_vec==1)
}

@@

patient_meta_data <- merge(clinical_data, patient_longitudinal_mapping_data, by.x="LAB_ID", by.y = "Overall_Lab_ID", all = TRUE)
patient_meta_data[ LAB_ID != LAB_ID.y]$LAB_ID <-  patient_meta_data[ LAB_ID != LAB_ID.y]$LAB_ID.y

@@

patient_meta_data %>%
  dplyr::filter(!is.na(manifestation_type)) %>%
  dplyr::select(Number.x, Sex, `Age at surgery (years)`, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`, `Mass Spectrometry`, Lipidomics, `RT-qPCR`, IHC, ELISA) %>%
  dplyr::group_by(Number.x) %>%
  dplyr::add_count() %>%
  unique() %>%
  dplyr::ungroup()-> meta_dt
```

@@

```{r echo=FALSE, fig.height=11, fig.width=21, message=FALSE, warning=FALSE}
## Data wrangling ----
### iGBM over rGBM ratio
data_long %>% 
  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
  dplyr::group_by(gene, Number.x) %>%
  dplyr::group_map(function(.d, ...){
    .d <- as.data.table(.d)
    .d %>% 
      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>% 
      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
    # print(.d)
    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
      .d_initial <- .d[manifestation_type=="Initial manifestation"]
      .d_second <- .d[manifestation_type=="Second manifestation"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else if (nrow(.d)==2) {
      .d_initial <- .d[collection_type=="E1"]
      .d_second <- .d[collection_type=="E2"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else {
      tmp$ratio <- .d$conc
    }
    return(tmp)
  }, .keep = TRUE) %>%
  rbindlist() %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
  dplyr::mutate(log2_conc = log2(ratio)) %>%
  # dplyr::group_by(gene, Number.x) %>%
  # dplyr::add_count() %>%
  # dplyr::filter(n==2) %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata

### rGBM over iGBM
data_long %>% 
  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
  dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
  dplyr::group_by(gene, Number.x) %>%
  dplyr::group_map(function(.d, ...){
    .d <- as.data.table(.d)
    .d %>% 
      dplyr::group_by(gene, Number.x, patient_state, `Age at surgery (years)`, Sex, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`) %>% 
      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
    # print(.d)
    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
      .d_initial <- .d[manifestation_type=="Initial manifestation"]
      .d_second <- .d[manifestation_type=="Second manifestation"]
      tmp$ratio <- .d_second$conc / .d_initial$conc
    } else if (nrow(.d)==2) {
      .d_initial <- .d[collection_type=="E1"]
      .d_second <- .d[collection_type=="E2"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else {
      tmp$ratio <- .d$conc
    }
    return(tmp)
  }, .keep = TRUE) %>%
  rbindlist() %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
  dplyr::filter(grepl(";", agg_lab_id)) %>%
  dplyr::mutate(log2_conc = log2(ratio)) %>%
  # dplyr::group_by(gene, Number.x) %>%
  # dplyr::add_count() %>%
  # dplyr::filter(n==2) %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2

@@

MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7, -8, -9, -10)], train_df$class, keepX = c(4,4)) 
background <- background.predict(MyResult.splsda, comp.predicted=2, dist = "max.dist")

plsda_data <- cbind( patient_id=na.omit(data_wide)  %>% dplyr::pull(Number.x), class=train_df$class, as.data.frame(MyResult.splsda$variates$X))
plsda_data$patient_id <- as.factor(plsda_data$patient_id)
plsda_data %>%
  dplyr::arrange(class) -> plsda_data
plsda_data <- as.data.table(plsda_data)

@@

>>>>

<!-- # Helper Functions -->
```{r message=FALSE, warning=FALSE, include=FALSE}
cs <- function(.x) {    # center/scale vector
  print(.x)
  out <- .x - mean(.x)  # center
  out / sd(out)         # scale
}

median_centre_data <- function(train_df){
  values_to = "measure_col"
  measure_col="measure_col"
  train_df %>%
  tibble::rowid_to_column(var = "sample_id_col") %>%
  tidyr::pivot_longer(cols = c(ASAH1, MMP.9, SYNEMIN, GPNMB), names_to = "feature_id_col", values_to = "measure_col") ->
  train_long

  normalized_df = train_long %>% group_by(sample_id_col) %>% 
      mutate(median_run = median(measure_col, na.rm = TRUE)) %>% 
      ungroup()
  old_measure_col = paste("preNorm", measure_col, sep = "_")
  normalized_df = normalized_df %>% mutate(median_global = median(!!(sym(measure_col)), 
    na.rm = TRUE), `:=`(!!(old_measure_col), !!(sym(measure_col)))) %>% 
    mutate(diff_norm = median_global - median_run) %>% mutate(`:=`(!!(sym(measure_col)), 
    !!(sym(measure_col)) + diff_norm))

  normalized_df = switch("default", all = normalized_df, 
      default = normalized_df, minimal = normalized_df %>% 
        dplyr::select(c(sample_id_col, feature_id_col, 
          measure_col, old_measure_col)))

  # ggplot(normalized_df, aes(x=as.factor(sample_id_col), y=measure_col, col=sample_id_col)) +
  #   geom_boxplot()

  normalized_df %>%
    dplyr::select(sample_id_col, class, feature_id_col, measure_col) %>%
    tidyr::pivot_wider(id_cols = c("sample_id_col", "class"), names_from = "feature_id_col", values_from = "measure_col") %>%
    tibble::column_to_rownames("sample_id_col") -> train_df

#' @title Master Trainer and Prediction
#' @description Top Function to perform predictions for testing logistic regression, logistic regression + elastic regularization, random forest and k-Nearest Neightbours
#' 
#' @param train_df (data.frame) A dataframe that contains labels and features to run through prediction models. Has to have the following structure:
#' \item{class}{(character) sample/observation label. i.e. ctrl, iGBM, or rGBM sample.}
#' \item{feature_1}{(numeric} a specific feature measurement such as protein ASAH1)}
#' \item{feature_n}{(numeric) another specific feature.}
#' @param pred_col_prob (character) A character corresponding to the class of probabilities you want to extract for. i.e. "GBM"
#' @param multiclass (logical) Run predictions using multiclass modelling and prediction if more than 2 classes.
#' 
#' @return
#' 
#' returns a list object that contains trained model objects, lift data and roc data.
#' 
#' \item(model){(glmnet) Logistic Regression with Elastic Net Regularization Model }
#' \item(model_rf){(rf) Random Forest Model}
#' \item(model_logit){(glm) Logistic Regression Model}
#' \item(model_boostedlogit){(knn) k-Nearest Neighbours Models}
#' \item(master_lift){(data.table) lift data for all models}
#' \item(roc_data){(data.table) roc data for all models}
run_predictions <- function(train_df, pred_col_prob="GBM", multiclass=FALSE){

@@

#' Calculate the ratio of iGBM to rGBM and rGBM to iGBM for a given data frame
#'
#' @param data_long2 A data frame with columns "gene", "Number.x", "patient_state", "Age at surgery (years)", "Sex", "Tumor localization", "Type of resection", "MGMT promotor methylation", "p53 accumulation", "Ki67-Li", "EGFR vIII", "Treatment", "Time-to-reccurence (days)", "Latency (days)", and "conc".
#' @return A data frame with the calculated ratios.
get_ratio_data <- function(data_long2){
  ## Data wrangling ----
  ### iGBM over rGBM ratio
  data_long2 %>% 
    dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
    dplyr::group_by(gene, Number.x) %>%
    dplyr::group_map(function(.d, ...){
      .d <- as.data.table(.d)
      .d %>% 
        dplyr::group_by(gene, Number.x, patient_state) %>% 
        dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                         agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
      # print(.d)
      if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
        .d_initial <- .d[manifestation_type=="Initial manifestation"]
        .d_second <- .d[manifestation_type=="Second manifestation"]
        tmp$ratio <- .d_initial$conc / .d_second$conc
      } else if (nrow(.d)==2) {
        .d_initial <- .d[collection_type=="E1"]
        .d_second <- .d[collection_type=="E2"]
        tmp$ratio <- .d_initial$conc / .d_second$conc
      } else {
        tmp$ratio <- NA #.d$conc
      }
      return(tmp)
    }, .keep = TRUE) %>%
    rbindlist() %>%
    dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
    dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
    dplyr::mutate(log2_conc = log2(ratio)) %>%
    # dplyr::group_by(gene, Number.x) %>%
    # dplyr::add_count() %>%
    # dplyr::filter(n==2) %>%
    dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata

  ### rGBM over iGBM
  data_long2 %>% 
    dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
    dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
    dplyr::group_by(gene, Number.x) %>%
    dplyr::group_map(function(.d, ...){
      .d <- as.data.table(.d)
      .d %>% 
        dplyr::group_by(gene, Number.x, patient_state) %>% 
        dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                         agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
      # print(.d)
      if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
        .d_initial <- .d[manifestation_type=="Initial manifestation"]
        .d_second <- .d[manifestation_type=="Second manifestation"]
        tmp$ratio <- .d_second$conc / .d_initial$conc
      } else if (nrow(.d)==2) {
        .d_initial <- .d[collection_type=="E1"]
        .d_second <- .d[collection_type=="E2"]
        tmp$ratio <- .d_second$conc / .d_initial$conc
      } else {
        tmp$ratio <- NA #.d$conc
      }
      return(tmp)
    }, .keep = TRUE) %>%
    rbindlist() %>%
    dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "post-surgery/pre-surgery"))) %>%
    dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
    dplyr::filter(grepl(";", agg_lab_id)) %>%
    dplyr::mutate(log2_conc = log2(ratio)) %>%
    # dplyr::group_by(gene, Number.x) %>%
    # dplyr::add_count() %>%
    # dplyr::filter(n==2) %>%
    dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2

@@

  ### Pivot Table to wide
  pdata %>%
    dplyr::select(Number.x, class, ASAH1, SYNEMIN, GPNMB, `MMP-9`) -> data_wide

@@

#' Calculate leave-one-patient-out cross-validation performance for splsda model
#'
#' This function takes in a dataset and a vector of patient IDs and returns a proportion
#' of correct predictions using a leave-one-patient-out cross-validation approach. A splsda
#' model is fit using the train data and used to make predictions on the test data. The
#' `unique_random_train_patient_per_class` parameter can be used to specify whether to 
#' include a unique patient for each "iGBM" and "rGBM" class in the train dataset. The
#' function returns a proportion of correct predictions.
#'
#' @param ratio_data A data frame containing the ratio data and class labels.
#' @param patient_ids A vector of patient IDs to use for the leave-one-patient-out
#' cross-validation.
#' @param unique_random_train_patient_per_class A logical value indicating whether to include
#' a unique patient for each "iGBM" and "rGBM" class in the train dataset.
#' @return A proportion of correct predictions.
#' @examples
#' splsda_loocv_perf(ratio_data, patient_ids)
splsda_loocv_perf <- function(ratio_data, patient_ids, unique_random_train_patient_per_class=TRUE){
  # Vector to store accuracy of LOOCV
  acc_vec <- numeric(length(patient_ids))
  i <- 1
  for (test_patient_id in patient_ids){
    # Split data into train and testing
    train_test <- get_train_test_for_loocv(ratio_data, test_patient_id=test_patient_id, unique_random_train_patient_per_class=unique_random_train_patient_per_class)
    train_df <- train_test$train_df
    test_df <- train_test$test_df
    # Perform splsda
    # MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7, -8, -9, -10)], train_df$class, keepX = c(4,4)) 
    MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2)], train_df$class, keepX = c(4,4)) 
    # Make prediction
    # pred_test <- predict(MyResult.splsda, test_df[, c(-1, -2, -7, -8, -9, -10)], test_df$class)
    pred_test <- predict(MyResult.splsda, test_df[, c(-1, -2)], test_df$class)
    # evaluate the prediction accuracy for the first two components
    predict.comp2 <- pred_test$class$mahalanobis.dist[,2]
    # compute confusion matrix
    conf_mat <- table(factor(predict.comp2, levels = levels(train_df$class)), test_df$class)
    # print(conf_mat)
    # compute accuracy
    acc <- (conf_mat[1,1] + conf_mat[2,2])/(conf_mat[1,1] + conf_mat[2,2] + conf_mat[1,2] + conf_mat[2,1])
    # message(sprintf("INFO: Test Patient %s - predicted class = %s - model accuracy = %s", test_patient_id, predict.comp2, acc))
    # Store accuracy
    acc_vec[i] <- acc
    # update index
    i <- i + 1
  }
  # Compute proportion of correct predictions
  mean(acc_vec==1)
}

splsda <- function(train_df, test_df){

    # Perform splsda
    MyResult.splsda <- mixOmics::splsda(train_df, train_df$class, keepX = c(4,4)) 
    # Make prediction
    pred_test <- predict(MyResult.splsda, test_df, test_df$class)
    # evaluate the prediction accuracy for the first two components
    predict.comp2 <- pred_test$class$mahalanobis.dist[,2]
    # compute confusion matrix
    conf_mat <- table(factor(predict.comp2, levels = levels(train_df$class)), test_df$class)
    # print(conf_mat)
    # compute accuracy
    acc <- (conf_mat[1,1] + conf_mat[2,2])/(conf_mat[1,1] + conf_mat[2,2] + conf_mat[1,2] + conf_mat[2,1])

    return(list(model=MyResult.splsda, acc=acc))
}
```

@@

patient_meta_data <- merge(clinical_data, patient_longitudinal_mapping_data, by.x="LAB_ID", by.y = "LAB_ID", all = TRUE)
# patient_meta_data[ LAB_ID != LAB_ID.y]$LAB_ID <-  patient_meta_data[ LAB_ID != LAB_ID.y]$LAB_ID.y

@@

patient_meta_data %>%
  dplyr::filter(!is.na(manifestation_type)) %>%
  dplyr::select(Number.x, Sex, `Age at surgery (years)`, `Tumor localization`, `Type of resection`, `MGMT promotor methylation`, `p53 accumulation`, `Ki67-Li`, `EGFR vIII`, Treatment, `Time-to-reccurence (days)`, `Latency (days)`, `Mass Spectrometry`, Lipidomics, `RT-qPCR`, IHC, ELISA) %>%
  dplyr::group_by(Number.x) %>%
  dplyr::add_count() %>%
  unique() %>%
  dplyr::ungroup()-> meta_dt

patient_id_meta <- unique(data_long[, .(LAB_ID, Number.x, collection_type, manifestation_type, patient_state, `Time-to-reccurence (days)`)])
```

@@

```{r echo=FALSE, fig.height=11, fig.width=21, message=FALSE, warning=FALSE}
## Data wrangling ----
### iGBM over rGBM ratio
data_long %>% 
  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
  dplyr::group_by(gene, Number.x) %>%
  dplyr::group_map(function(.d, ...){
    .d <- as.data.table(.d)
    .d %>% 
      dplyr::group_by(gene, Number.x, patient_state) %>% 
      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
    # print(.d)
    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
      .d_initial <- .d[manifestation_type=="Initial manifestation"]
      .d_second <- .d[manifestation_type=="Second manifestation"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else if (nrow(.d)==2) {
      .d_initial <- .d[collection_type=="E1"]
      .d_second <- .d[collection_type=="E2"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else {
      tmp$ratio <- .d$conc
    }
    return(tmp)
  }, .keep = TRUE) %>%
  rbindlist() %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "iGBMoverrGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="iGBMoverrGBM" ) %>%
  dplyr::mutate(log2_conc = log2(ratio)) %>%
  # dplyr::group_by(gene, Number.x) %>%
  # dplyr::add_count() %>%
  # dplyr::filter(n==2) %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("iGBMoverrGBM"))) -> pdata

### rGBM over iGBM
data_long %>% 
  dplyr::mutate(conc = conc+1) %>% # Add one to avoid cases of 0 division
  dplyr::filter(!is.na(manifestation_type) | patient_state!=0) %>%
  dplyr::group_by(gene, Number.x) %>%
  dplyr::group_map(function(.d, ...){
    .d <- as.data.table(.d)
    .d %>% 
      dplyr::group_by(gene, Number.x, patient_state) %>% 
      dplyr::summarise(agg_lab_id = paste(LAB_ID, collapse=";"),
                       agg_collection_type = paste(collection_type, collapse = ";"), .groups = "keep") -> tmp
    # print(.d)
    if ( nrow(.d)==2 & any(!is.na(.d$manifestation_type)) ){
      .d_initial <- .d[manifestation_type=="Initial manifestation"]
      .d_second <- .d[manifestation_type=="Second manifestation"]
      tmp$ratio <- .d_second$conc / .d_initial$conc
    } else if (nrow(.d)==2) {
      .d_initial <- .d[collection_type=="E1"]
      .d_second <- .d[collection_type=="E2"]
      tmp$ratio <- .d_initial$conc / .d_second$conc
    } else {
      tmp$ratio <- .d$conc
    }
    return(tmp)
  }, .keep = TRUE) %>%
  rbindlist() %>%
  dplyr::mutate(sample_comparison = plyr::mapvalues(patient_state, from=c(0, 1, 2), to=c("ctrl", "rGBMoveriGBM", "pre-surgery/post-surgery"))) %>%
  dplyr::filter(sample_comparison=="rGBMoveriGBM" ) %>%
  dplyr::filter(grepl(";", agg_lab_id)) %>%
  dplyr::mutate(log2_conc = log2(ratio)) %>%
  # dplyr::group_by(gene, Number.x) %>%
  # dplyr::add_count() %>%
  # dplyr::filter(n==2) %>%
  dplyr::mutate(class = plyr::mapvalues(patient_state, from=c(1), to=c("rGBMoveriGBM"))) -> pdata2

@@

org_class <- train_df$class
train_df$class <- sample(train_df$class)

MyResult.splsda <- mixOmics::splsda(train_df[, c(-1, -2, -7, -8, -9, -10)], train_df$class, keepX = c(4,4)) 
MyResult.splsda <- mixOmics::splsda(train_df[, c(-1)], train_df$class, keepX = c(4,4))
background <- background.predict(MyResult.splsda, comp.predicted=2, dist = "max.dist")

plsda_data <- cbind( patient_id=na.omit(data_wide %>% dplyr::filter(Number.x!=32))  %>% dplyr::pull(Number.x), class=train_df$class, as.data.frame(MyResult.splsda$variates$X))
plsda_data$patient_id <- as.factor(plsda_data$patient_id)
plsda_data %>%
  dplyr::arrange(class) -> plsda_data
plsda_data <- as.data.table(plsda_data)

@@

background %>%
  dplyr::filter(Var2 < 1.59058 & Var2 > -1.56044 & Var1 > -2.5462 & Var1 < 2.36565) -> background_line

lm_eq <- lm(Var2 ~ Var1, data=background_line)
coef(lm_eq)

plsda_data %>%
  dplyr::rowwise() %>%
  dplyr::mutate(new_x = (comp1 + -1/coef(lm_eq)[2]*comp2  - (-1/coef(lm_eq)[2]*(-coef(lm_eq)[1])))/(1+(-1/coef(lm_eq)[2])^2),
                new_y = (-1/coef(lm_eq)[2]*comp1 + ((-1/coef(lm_eq)[2])^2)*comp2  + (-coef(lm_eq)[1]))/(1+(-1/coef(lm_eq)[2])^2)) -> plsda_data_proj

ggplot() +
  # geom_point(data=background_line, aes(x=Var1, y=Var2)) +
  geom_abline(slope = coef(lm_eq)[2], intercept = coef(lm_eq)[1]) +
  geom_abline(slope = -1/coef(lm_eq)[2], intercept = -coef(lm_eq)[1]) +
  geom_point(data=plsda_data_proj, aes(x=new_x, y=new_y, col=class))
